Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2022-08-19 manuscript.Rmd
small updates

  This paper is written in R Markdown. The R Markdown file, other R code, and additional information on this study can be found in the [Online Supplement](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

If estimated for each participant independently, observed individual effects are quite variable because they are perturbed by sample noise. In hierarchical modeling, these individual effects are optimally and automatically corrected towards the general trend [@efron1977stein]. This phenomenon, also called "shrinkage", is illustrated in Panel B of Figure \@ref(fig:figure1). The rightmost observation in the figure is in sample quite large compared to the other observed effects. The estimated effect (green point) is drastically corrected towards the mean. Conversely, the general trend is only influenced slightly by individuals showing a highly divergent trend. By adding the individual variability to the model, we no longer assume independence between observations, but the overall effect in the model leaves the possibility to generalize the effect. Therefore, hierarchical analysis can be considered a middle ground between an independent analysis and an analysis of aggregated data. 
```{r figure2, echo=FALSE, eval = FALSE, fig.cap="Sample represents the observed response time in milliseconds per individual, while model represents the individual response time estimates. Individual estimates are shrunken towards the general trend. ", fig.align = "left", fig.height = 2, fig.width = 3, out.width="70%"}
knitr::include_graphics("Images/hierplotefmor.png")
```

@@

```{r hierarchyplot, echo=FALSE, fig.cap="Hierarchical structure of the normal and lognormal models for the digit classification task.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="70%"}
knitr::include_graphics("Images/Fig2_Hierarchical structure.png")
```

@@

```{r priorpredfigurepanel, echo=FALSE, fig.cap="Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \\textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the lognormal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the lognormal model based on the selected priors.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%"}
knitr::include_graphics("Images/Fig5_priorpredfigure_v2.png")
```

@@

## Software 
There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used `R` packages, `rstan` [@R-rstan] and `brms` [@R-brms_a; @R-brms_b]. We assume that readers are somewhat familiar with `R`. The package `rstan` provides the most modern algorithms and flexibility in model set-up. The package `brms` was created as an overlay to `rstan` and is based on the well-known `lme4` syntax [@R-lme4]. The goal of `brms` is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer a separate Rmarkdown file for each package in our [Online Supplement](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment) that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

# Estimation of the Digit Classification Task
## Data 
For our example, we will use the data set by Rouder et al. (2005). The data are available under a free license on GitHub [https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5](https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5). The data set contains observations from 54 participants who took part in a digit classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (< 250 MS) and very slow RTs (> 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by @haaf2019capturing. This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

### Estimation Results
Considering good/acceptable convergence diagnostics, we can now interpret the estimation results. First, we present the general effects, then we visualize the individual variability of these effects.

@@

We do not have ownership of the data used in this study, but the the data set is publicly available on [GitHub](https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/). The `.stan` models, results of the model fit, and model comparison are available on [our GitHub repository](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

The R code used for the analyses and figures in this manuscript, and `R` tutorials on BHMs are available on [our GitHub repository](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

>>>>

  This paper is written in R Markdown. The R Markdown file, other R code, and additional information on this study can be found in the [Online Supplement: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

If estimated for each participant independently, observed individual effects are quite variable because they are perturbed by sample noise. In hierarchical modeling, these individual effects are optimally and automatically corrected towards the general trend [@efron1977stein]. This phenomenon, also called `shrinkage', is illustrated in Panel B of Figure \@ref(fig:figure1). The rightmost observation in the figure is in sample quite large compared to the other observed effects. The estimated effect (green point) is drastically corrected towards the mean. Conversely, the general trend is only influenced slightly by individuals showing a highly divergent trend. By adding the individual variability to the model, we no longer assume independence between observations, but the overall effect in the model leaves the possibility to generalize the effect. Therefore, hierarchical analysis can be considered a middle ground between an independent analysis and an analysis of aggregated data. 
```{r figure2, echo=FALSE, eval = FALSE, fig.cap="Sample represents the observed response time in milliseconds per individual, while model represents the individual response time estimates. Individual estimates are shrunken towards the general trend. ", fig.align = "left", fig.height = 2, fig.width = 3, out.width="70%"}
knitr::include_graphics("Images/hierplotefmor.png")
```

@@

```{r hierarchyplot, echo=FALSE, fig.cap="Hierarchical structure of the normal and log-normal models for the digit classification task.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="70%"}
knitr::include_graphics("Images/Fig2_Hierarchical structure.png")
```

@@

```{r priorpredfigurepanel, echo=FALSE, fig.cap="Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \\textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the log-normal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the log-normal model based on the selected priors.", fig.align = "left", fig.height = 3, fig.width = 4, out.width="100%"}
knitr::include_graphics("Images/Fig5_priorpredfigure_v2.png")
```

@@

## Software 
There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used `R` packages, `rstan` [@R-rstan] and `brms` [@R-brms_a; @R-brms_b]. We assume that readers are somewhat familiar with `R`. The package `rstan` provides the most modern algorithms and flexibility in model set-up. The package `brms` was created as an overlay to `rstan` and is based on the well-known `lme4` syntax [@R-lme4]. The goal of `brms` is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer a separate R Markdown file for each package in our [Online Supplement](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment) that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

# Estimation of the Digit Classification Task
## Data 
For our example, we will use the data set by Rouder et al. (2005). The data are available under a free license on GitHub [https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5]((https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5)). The data set contains observations from 54 participants who took part in a digit classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (< 250 MS) and very slow RTs (> 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by @haaf2019capturing. This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

### Estimation Results
Considering acceptable to good convergence diagnostics, we can now interpret the estimation results. First, we present the general effects, then we visualize the individual variability of these effects.

@@

We do not have ownership of the data used in this study, but the the data set is publicly available on [GitHub (https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/)](https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/). The `.stan` models, results of the model fit, and model comparison are available on [our GitHub repository: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

The R code used for the analyses and figures in this manuscript, and `R` tutorials on BHMs are available on [our GitHub repository: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.2in}

### 2022-08-19 manuscript.tex
small updates

This paper is written in R Markdown. The R Markdown file, other R code, and additional information on this study can be found in the \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement}.

@@

If estimated for each participant independently, observed individual effects are quite variable because they are perturbed by sample noise. In hierarchical modeling, these individual effects are optimally and automatically corrected towards the general trend (Efron \& Morris, 1977). This phenomenon, also called ``shrinkage,'' is illustrated in Panel B of Figure \ref{fig:figure1}. The rightmost observation in the figure is in sample quite large compared to the other observed effects. The estimated effect (green point) is drastically corrected towards the mean. Conversely, the general trend is only influenced slightly by individuals showing a highly divergent trend. By adding the individual variability to the model, we no longer assume independence between observations, but the overall effect in the model leaves the possibility to generalize the effect. Therefore, hierarchical analysis can be considered a middle ground between an independent analysis and an analysis of aggregated data.

@@

\caption{Hierarchical structure of the normal and lognormal models for the digit classification task.}\label{fig:hierarchyplot}
\end{figure}

@@

\caption{Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the lognormal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the lognormal model based on the selected priors.}\label{fig:priorpredfigurepanel}
\end{figure}

@@

There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used \texttt{R} packages, \texttt{rstan} (Stan Development Team, 2019a) and \texttt{brms} (Bürkner, 2017, 2018). We assume that readers are somewhat familiar with \texttt{R}. The package \texttt{rstan} provides the most modern algorithms and flexibility in model set-up. The package \texttt{brms} was created as an overlay to \texttt{rstan} and is based on the well-known \texttt{lme4} syntax (Bates, Mächler, Bolker, \& Walker, 2015). The goal of \texttt{brms} is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer a separate Rmarkdown file for each package in our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement} that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

For our example, we will use the data set by Rouder et al.~(2005). The data are available under a free license on GitHub \url{https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5}. The data set contains observations from 54 participants who took part in a digit classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (\textless{} 250 MS) and very slow RTs (\textgreater{} 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by Haaf, Klaassen, and Rouder (2019). This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

Considering good/acceptable convergence diagnostics, we can now interpret the estimation results. First, we present the general effects, then we visualize the individual variability of these effects.

@@

We do not have ownership of the data used in this study, but the the data set is publicly available on \href{https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/}{GitHub}. The \texttt{.stan} models, results of the model fit, and model comparison are available on \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{our GitHub repository}.

@@

The R code used for the analyses and figures in this manuscript, and \texttt{R} tutorials on BHMs are available on \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{our GitHub repository}.

>>>>

This paper is written in R Markdown. The R Markdown file, other R code, and additional information on this study can be found in the \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}.

@@

If estimated for each participant independently, observed individual effects are quite variable because they are perturbed by sample noise. In hierarchical modeling, these individual effects are optimally and automatically corrected towards the general trend (Efron \& Morris, 1977). This phenomenon, also called `shrinkage', is illustrated in Panel B of Figure \ref{fig:figure1}. The rightmost observation in the figure is in sample quite large compared to the other observed effects. The estimated effect (green point) is drastically corrected towards the mean. Conversely, the general trend is only influenced slightly by individuals showing a highly divergent trend. By adding the individual variability to the model, we no longer assume independence between observations, but the overall effect in the model leaves the possibility to generalize the effect. Therefore, hierarchical analysis can be considered a middle ground between an independent analysis and an analysis of aggregated data.

@@

\caption{Hierarchical structure of the normal and log-normal models for the digit classification task.}\label{fig:hierarchyplot}
\end{figure}

@@

\caption{Side and digit effects from predicted data. Violins represent the distribution of predicted effects across 1000 simulation runs. The blue points are the average effects across simultions. Panel A shows the prior predictions for the normal model based on the default priors of \textit{brms}. Panel B shows the prior predictions for the normal model based on the selected priors. Panel C shows the prior predictions for the log-normal model based on the priors chosen for the normal model. Panel D shows the prior predictions for the log-normal model based on the selected priors.}\label{fig:priorpredfigurepanel}
\end{figure}

@@

There are many different programs available to fit Bayesian hierarchical models, that is, to obtain samples from the joint posterior distribution of parameters. In this tutorial, we focus on two of the most commonly used \texttt{R} packages, \texttt{rstan} (Stan Development Team, 2019a) and \texttt{brms} (Bürkner, 2017, 2018). We assume that readers are somewhat familiar with \texttt{R}. The package \texttt{rstan} provides the most modern algorithms and flexibility in model set-up. The package \texttt{brms} was created as an overlay to \texttt{rstan} and is based on the well-known \texttt{lme4} syntax (Bates, Mächler, Bolker, \& Walker, 2015). The goal of \texttt{brms} is to ease the transition to Bayesian hierarchical modeling for novice users of Bayesian statistics. The packages work similarly for the normal and log-normal model. In this section, we explain how a model can be fitted in each package using code snippets. In addition, we offer a separate R Markdown file for each package in our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{Online Supplement} that explains in more depth how to analyze the full symbolic distance effect model with the specific package.

@@

For our example, we will use the data set by Rouder et al.~(2005). The data are available under a free license on GitHub \href{(https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5)}{https://github.com/PerceptionCognitionLab/data0/tree/master/lexDec-dist5}. The data set contains observations from 54 participants who took part in a digit classification task. Each participant performed 360 trials in six blocks. For the analysis, only the correct responses are included. To account for practice effects, the first 25 trials of the first block are removed from the analysis. The trials after the breaks between the blocks are also removed from the analysis to correct for lack of concentration. In addition, trials with very fast RTs (\textless{} 250 MS) and very slow RTs (\textgreater{} 2000 MS) are excluded. Two participants were excluded because they did not respond properly to the task. Similar criteria were applied by Haaf, Klaassen, and Rouder (2019). This results in 52 participants and a total of 17031 observations. In the following two sections we will describe the model fitting and parameter estimation procedure for the normal and log-normal model, respectively.

@@

Considering acceptable to good convergence diagnostics, we can now interpret the estimation results. First, we present the general effects, then we visualize the individual variability of these effects.

@@

We do not have ownership of the data used in this study, but the the data set is publicly available on \href{https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/}{GitHub (https://github.com/PerceptionCognitionLab/data0/blob/master/lexDec-dist5/)}. The \texttt{.stan} models, results of the model fit, and model comparison are available on \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{our GitHub repository: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}.

@@

The R code used for the analyses and figures in this manuscript, and \texttt{R} tutorials on BHMs are available on \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}{our GitHub repository: https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment}.
